Rectangulotopes

### 2024-04-05 draft/rectangulotopes.tex
more pictures Section 2

The encoding of combinatorial objects in the form of a polyhedral structure is a recurrent theme in geometric and algebraic combinatorics.
An elementary yet striking example of this idea is the \defn{permutahedron} $\Perm (n)$, defined as the convex hull of the points $\sum_{i\in [n]} \sigma^{-1}(i)\cdot \b{e}_i$, for all permutations $\sigma \in \f{S}_n$.
It is an $(n-1)$-dimensional polytope, whose facets are indexed by the proper nonempty subsets of $[n]$, and whose skeleton is the Cayley graph of the symmetric group for the generators consisting of adjacent transpositions.
The many generalizations and extensions of the permutahedra gave rise to a flourishing theory of \defn{generalized permutahedra}~\cite{Postnikov} (that we prefer to call \defn{deformed permutahedra}).

Among those, the \defn{associahedron} is a classical, ubiquitous polytope, first defined combinatorially by D.~Tamari~\cite{T51} and by J.~Stasheff in a topological context~\cite{S63}.
It has since then found applications in various areas of mathematics, including topology, operads, cluster algebras, combinatorial Hopf algebras, and physics~\cite{PilaudSantosZiegler}.
It was first defined as a combinatorial object, before various families of geometric realizations were shown to exist~\cite{MR1022776,MR1941227,MR2108555,MR3437894}.
The face lattice of the $(n-1)$-dimensional associahedron is the reverse inclusion poset of nonintersecting diagonals of a convex $(n+3)$-gon.
In particular, its skeleton is the \defn{flip graph} on \defn{triangulations} of the $(n+3)$-gon, or equivalently -- via a standard Catalan bijection -- the \defn{rotation graph} on \defn{binary trees} with $n$ internal nodes, the structure of which has been the subject of many investigations, with applications in computer science~\cite{MR928904,MR3197650}.
In 2004, J.-L.~Loday published the following elegant description of the associahedron~\cite{MR2108555}, giving a recipe for computing the coordinates of a vertex given the corresponding binary tree.
We note that the inequality description of the same polytope was actually provided in 1993 by S.~Shnider and S.~Sternberg~\cite{ShniderSternberg}, but Loday's vertex description largely popularized this realization.

@@

A natural object associated with the associahedron is the \defn{Tamari lattice}, the cover graph of which is isomorphic to the skeleton of the associahedron.
The Tamari lattice is known to be the quotient of the weak Bruhat order by a \defn{lattice congruence} relation called the \defn{sylvester congruence}.
Answering a question of N.~Reading~\cite{MR2142177}, V.~Pilaud and F.~Santos~\cite{MR3964495} proved that with \emph{every} lattice congruence of the weak Bruhat order, one can associate a polytope whose skeleton is the cover graph of the lattice quotient, and more precisely whose normal fan is the \defn{quotient fan} defined by gluing the cones of the \defn{braid fan} (the type $A$ Coxeter arrangement) that belong to the same congruence class.
Those polytopes are deformed permutahedra that they called \defn{quotientopes}.
Associahedra are therefore the quotientopes of the sylvester congruence of the weak Bruhat order whose classes are in bijection with triangulations.

The goal of this paper is to describe the quotientopes of congruences of the weak Bruhat order whose classes are in bijection with equivalence classes of rectangulations, defined as decompositions of a square into rectangles.

@@

\begin{figure}
  \caption{\label{fig:rectequiv}Weak and strong equivalence between rectangulations.}
\end{figure}

@@

The \defn{weak rectangulation congruence} (or \defn{Baxter congruence}) was first explicitly studied by S.~Law and N.~Reading~\cite{MR2871762} and revisited in~\cite[Thm.~1.1 \& Exm.~4.10]{Reading-arcDiagrams}.
Its classes are in bijection with with weak rectangulations, but also with \defn{twin binary trees}~\cite{Giraudo}, and with \defn{Baxter permutations}~\cite{}. % more baxter refs
The corresponding quotientopes will be referred to as the \defn{weak rectangulotopes} and denoted by $\WRP(n)$.
%The skeleton of $\WRP(n)$ is isomorphic to the flip graph of weak rectangulations of size $n$ defined by Law and Reading~\cite{MR2871762}.

The \defn{strong rectangulation congruence} was introduced by N.~Reading~\cite{MR2864445}, revisited in~\cite[Thm.~1.2 \& Exm.~4.11]{Reading-arcDiagrams}, and studied more recently by A.~Asinowski, J.~Cardinal, S.~Felsner, and É.~Fusy~\cite{ACFF24}.
Its classes are in bijection with strong rectangulations.
The corresponding quotientopes will be referred to as the \defn{strong rectangulotopes} and denoted by $\SRP(n)$.

@@

\begin{lemma}
%  The skeleton of $\WRP(n)$ is isomorphic to the flip graph on weak rectangulations of size~$n$.
%  Similarly, the skeleton of $\SRP(n)$ is isomorphic to the flip graph on strong rectangulations of size~$n$.
  The skeleton of $\WRP(n)$ (resp.~of $\SRP(n)$) is isomorphic to the flip graph on weak (resp.~strong) rectangulations of size~$n$.
\end{lemma}

@@

\parpic(4cm,2.5cm)(10pt, 150pt)[r][b]{\includegraphics[scale=.9]{arcDiagram}}{
In~\cite{Reading-arcDiagrams}, N.~Reading defined an elegant bijection between permutations of~$[n]$ and noncrossing arc diagrams on~$[n]$.
It sends a permutation~$\sigma$ to the noncrossing arc diagram with an arc~$(\sigma_{j+1}, \sigma_j, A_j, B_j)$ for each descent~$j \in [n-1]$ of~$\sigma$ (\ie with~$\sigma_j > \sigma_{j+1}$), where

@@

For instance, the \defn{sylvester congruence} is ... \vincent{todo}
See \cref{fig:sylvesterCongruence}\,(middle).
The quotient of the weak Bruhat order by the sylvester congruence is the classical \defn{Tamari lattice}.
See \cref{fig:sylvesterCongruence}\,(right).

@@

%Moreover, the sets of arcs of the form~$\c{A}_\equiv$ are easily described.
An arc~$(a, b, A, B)$ is a \defn{subarc} of an arc~$(a', b', A', B')$ if~$a' \le a < b \le b$ and~$A \subseteq A'$ while~$B \subseteq B'$.
An \defn{arc ideal} is a subset of arcs closed by subarcs.
The map~${\equiv} \mapsto \c{A}_\equiv$ is a bijection between the lattice congruences of the weak Bruhat order and the arc ideals.
In other words, the lattice of congruences of the weak Bruhat order is distributive, and its poset of join irreducibles is isomorphic to the subarc order.

@@

\begin{figure}
	\capstart
	\centerline{\includegraphics[scale=.6]{braidFanLeft4} \qquad \raisebox{-.3cm}{\includegraphics[scale=.6]{sylvesterFan4}}}
	\caption{The braid fan (left) and the sylvester fan (right). \cite[Fig.~1]{PilaudSantos-quotientopes}}
	\label{fig:fans}
\end{figure}

The \defn{braid arrangement} is the hyperplane arrangement consisting of the hyperplanes~$\set{\b{x} \in \R^n}{x_a = x_b}$ for all~$1 \le a < b \le n$.
It has a chamber for each permutation~$\sigma$ of~$[n]$, given by the set of vectors whose coordinates are ordered as~$\sigma$.

The \defn{shard} of an arc~$(a, b, A, B)$ is the piece of the braid hyperplane~${x_a = x_b}$ defined by the inequalities~$x_{a'} < x_a$ for all~$a' \in A$ and~$x_b < x_{b'}$ for all~$b' \in B$.
%$\Sigma(a, b, A, B) \eqdef \set{\b{x} \in \R^n}{x_a = x_b \text{ and } x_{a'} < x_a \text{ for all } a' \in A \text{ and }  x_b < x_{b'} \text{ for all } b' \in B}$.

The \defn{quotient fan} of a lattice congruence~$\equiv$ of the weak Bruhat order is the polyhedral fan~$\c{F}_\equiv$ where
\begin{itemize}
\item the maximal cones are obtained by glueing together the chambers of the braid arrangement corresponding to permutations in the same congruence class of~$\equiv$,
\item the union of the codimension~$1$ cones is the union of the shards of the arcs of~$\c{A}_\equiv$.
\end{itemize}
(These two descriptions are equivalent.)
By construction, the braid fan refines the quotient fan~$\c{F}_\equiv$, and the dual graph of the quotient fan~$\c{F}_\equiv$ is isomorphic to the cover graph of the quotient~$\f{S}_n/{\equiv}$.

@@

\begin{figure}
	\capstart
	\centerline{\includegraphics[scale=.6]{permutahedronLeft4} \qquad \raisebox{-.4cm}{\includegraphics[scale=.6]{associahedron4}}}
	\caption{The permutahedron~$\Perm(4)$ (left) and the associahedron~$\Asso(4)$. \cite[Fig.~1]{PilaudSantos-quotientopes}}
	\label{fig:quotientopes}
\end{figure}

A \defn{quotientope} for a lattice congruence~$\equiv$ of the weak Bruhat order is a polytope whose normal fan is the quotient fan~$\c{F}_\equiv$.
In particular, the skeleton of a quotientope is isomorphic to the cover graph of the quotient~$\f{S}_n/{\equiv}$.
The existence of quotientopes was first proved in \cite{PilaudSantos} and later better understood in~\cite{PadrolPilaudRitter} using Minkowski sums of shard polytopes.

The \defn{Minkowski sum} of two polytopes~$\polytope{P}$ and~$\polytope{Q}$ is~$\polytope{P} + \polytope{Q} \eqdef \set{\b{p} + \b{q}}{\b{p} \in \polytope{P} \text{ and } \b{q} \in \polytope{Q}}$.
Recall that for any vector~$\b{v} \ne \b{0}$, the face of~$\polytope{P} + \polytope{Q}$ maximizing the scalar product with~$\b{v}$ is the Minkowski sum of the faces of~$\polytope{P}$ and~$\polytope{Q}$ maximizing the scalar product with~$\b{v}$.
In particular, if~$\b{v}$ is a generic direction, the vertex of~$\polytope{P} + \polytope{Q}$ extremal in direction~$\b{v}$ is just the sum of the vertices of~$\polytope{P}$ and~$\polytope{Q}$ that are extremal in direction~$\b{v}$.

@@

@@

\begin{lemma}
  \label{lem:submodsum}
  Consider a collection $\polytope{P}_1, \polytope{P}_2,\ldots ,\polytope{P}_k$ of deformed permutahedra defined by the submodular functions $f_1,f_2,\ldots ,f_k$. Then the Minkowski sum $\sum_i \polytope{P}_i$ is defined by the submodular function $f \eqdef \sum_i f_i$.
\end{lemma}

@@

The description of the weak rectangulotope~$\WRP(n)$ as a Minkowski sum in \cref{lem:weakMinkowski} allows us to easily compute the corresponding submodular function.

>>>>

The encoding of combinatorial objects in the form of a polyhedral structure is a recurrent theme in geometric and algebraic combinatorics.
An elementary yet striking example of this idea is the \defn{permutahedron} $\Perm (n)$, defined as the convex hull of the points $\sum_{i\in [n]} \sigma^{-1}(i)\cdot \b{e}_i$, for all permutations $\sigma \in \f{S}_n$.
It is an $(n-1)$-dimensional polytope, whose facets are indexed by the proper nonempty subsets of $[n]$, and whose skeleton is the Cayley graph of the symmetric group for the generators consisting of adjacent transpositions.
See \cref{fig:quotientopes}\,(left) for an illustration when~$n = 4$.
The many generalizations and extensions of the permutahedra gave rise to a flourishing theory of \defn{generalized permutahedra}~\cite{Postnikov} (that we prefer to call \defn{deformed permutahedra}).

Among those, the \defn{associahedron} is a classical, ubiquitous polytope, first defined combinatorially by D.~Tamari~\cite{T51} and by J.~Stasheff in a topological context~\cite{S63}.
It has since then found applications in various areas of mathematics, including topology, operads, cluster algebras, combinatorial Hopf algebras, and physics~\cite{PilaudSantosZiegler}.
It was first defined as a combinatorial object, before various families of geometric realizations were shown to exist~\cite{MR1022776,MR1941227,MR2108555,MR3437894}.
The face lattice of the $(n-1)$-dimensional associahedron is the reverse inclusion poset of nonintersecting diagonals of a convex $(n+3)$-gon.
In particular, its skeleton is the \defn{flip graph} on \defn{triangulations} of the $(n+3)$-gon, or equivalently -- via a standard Catalan bijection -- the \defn{rotation graph} on \defn{binary trees} with $n$ internal nodes, the structure of which has been the subject of many investigations, with applications in computer science~\cite{MR928904,MR3197650}.
In 2004, J.-L.~Loday published the following elegant description of the associahedron~\cite{MR2108555}, giving a recipe for computing the coordinates of a vertex given the corresponding binary tree.
We note that the inequality description of the same polytope was actually provided in 1993 by S.~Shnider and S.~Sternberg~\cite{ShniderSternberg}, but Loday's vertex description largely popularized this realization.
See \cref{fig:quotientopes}\,(right) for an illustration when~$n = 4$.

@@

A natural object associated with the associahedron is the \defn{Tamari lattice}, whose cover graph is isomorphic to the skeleton of the associahedron.
The Tamari lattice is known to be the quotient of the weak Bruhat order by a \defn{lattice congruence} called the \defn{sylvester congruence}.
Answering a question of N.~Reading~\cite{MR2142177}, V.~Pilaud and F.~Santos~\cite{MR3964495} proved that with \emph{every} lattice congruence of the weak Bruhat order, one can associate a polytope whose skeleton is the cover graph of the lattice quotient, and more precisely whose normal fan is the \defn{quotient fan} defined by gluing the cones of the \defn{braid fan} (the type $A$ Coxeter arrangement) that belong to the same congruence class.
Those polytopes are deformed permutahedra that they called \defn{quotientopes}.
Associahedra are therefore the quotientopes of the sylvester congruence of the weak Bruhat order whose classes are in bijection with triangulations.

The goal of this paper is to describe the geometry of the quotientopes for two particular congruences of the weak Bruhat order whose classes are in bijection with equivalence classes of rectangulations, defined as decompositions of a square into rectangles.

@@

\begin{figure}
	\caption{Weak and strong equivalence between rectangulations.}
	\label{fig:rectequiv}
\end{figure}

@@

The \defn{weak rectangulation congruence} (or \defn{Baxter congruence}) was first explicitly studied by S.~Law and N.~Reading~\cite{MR2871762} and revisited in~\cite[Thm.~1.1 \& Exm.~4.10]{Reading-arcDiagrams} in terms of arc diagrams.
Its classes are in bijection with weak rectangulations, but also with \defn{twin binary trees}~\cite{Giraudo}, and with \defn{Baxter permutations}~\cite{}. % more baxter refs
The corresponding quotientopes will be referred to as the \defn{weak rectangulotopes} and denoted by~$\WRP(n)$.
%The skeleton of $\WRP(n)$ is isomorphic to the flip graph of weak rectangulations of size $n$ defined by Law and Reading~\cite{MR2871762}.

The \defn{strong rectangulation congruence} was introduced by N.~Reading~\cite{MR2864445}, revisited in~\cite[Thm.~1.2 \& Exm.~4.11]{Reading-arcDiagrams} in terms of arc diagrams, and studied more recently by A.~Asinowski, J.~Cardinal, S.~Felsner, and É.~Fusy~\cite{ACFF24}.
Its classes are in bijection with strong rectangulations.
The corresponding quotientopes will be referred to as the \defn{strong rectangulotopes} and denoted by~$\SRP(n)$.

@@

\begin{figure}
	\caption{Three types of flips between rectangulations.}
	\label{fig:flip}
\end{figure}

\begin{lemma}
%  The skeleton of $\WRP(n)$ is isomorphic to the flip graph on weak rectangulations of size~$n$.
%  Similarly, the skeleton of $\SRP(n)$ is isomorphic to the flip graph on strong rectangulations of size~$n$.
  The skeleton of $\WRP(n)$ (resp.~of $\SRP(n)$) is isomorphic to the flip graph on weak (resp.~strong) rectangulations of size~$n$.
  \vincent{I would leave that in the text, not in a proper thm environment...}
\end{lemma}

@@

\vspace{-.1cm}
\parpic(4cm,2.5cm)(10pt, 150pt)[r][b]{\includegraphics[scale=.9]{arcDiagram}}{
In~\cite{Reading-arcDiagrams}, N.~Reading defined an elegant bijection between permutations of~$[n]$ and noncrossing arc diagrams on~$[n]$.
It sends a permutation~$\sigma$ to the noncrossing arc diagram with an arc~$(\sigma_{j+1}, \sigma_j, A_j, B_j)$ for each descent~$j \in [n-1]$ of~$\sigma$ (\ie with~$\sigma_j > \sigma_{j+1}$), where

@@

For instance, the sets of linear extensions of binary trees (labeled in inorder and oriented toward their roots) are the classes of the \defn{sylvester congruence}~\cite{HivertNovelliThibon-binarySearchTrees}.
See \cref{fig:sylvesterCongruence}\,(middle).
The quotient of the weak Bruhat order by the sylvester congruence is the classical \defn{Tamari lattice}.
See \cref{fig:sylvesterCongruence}\,(right).

@@

%Moreover, the sets of arcs of the form~$\c{A}_\equiv$ are easily described.
\vspace{-.15cm}
\parpic(5cm,2.8cm)(10pt, 90pt)[r][b]{\includegraphics[scale=.5]{subarcOrder}}{
An arc~$(a, b, A, B)$ is a \defn{subarc} of an arc~$(a', b', A', B')$ if \linebreak $a' \le a < b \le b$ and~$A \subseteq A'$ while~$B \subseteq B'$.
An \defn{arc ideal} is a subset of arcs closed by subarcs.
The map~${\equiv} \mapsto \c{A}_\equiv$ is a bijection between the lattice congruences of the weak Bruhat order and the arc ideals.
In other words, the lattice of congruences of the weak Bruhat order is distributive, and its poset of join irreducibles is isomorphic to the subarc order, illustrated on the right for~$n = 4$.
}

@@

\begin{figure}[b]
	\capstart
	\centerline{\includegraphics[scale=.6]{braidFanLeft4} \qquad \raisebox{-.3cm}{\includegraphics[scale=.6]{sylvesterFan4}}}
	\caption{The braid fan (left) and the sylvester fan (right). \cite[Fig.~1]{PilaudSantos-quotientopes} \&~\cite[Fig.~5]{PadrolPilaudRitter}}
	\label{fig:fans}
\end{figure}

The \defn{braid arrangement} is the hyperplane arrangement consisting of the hyperplanes~$\set{\b{x} \in \R^n}{x_a = x_b}$ for all~$1 \le a < b \le n$.
It has a chamber for each permutation~$\sigma$ of~$[n]$, given by the set of vectors whose coordinates are ordered as~$\sigma$.
See \cref{fig:fans}\,(left).

The \defn{shard} of an arc~$(a, b, A, B)$ is the piece of the braid hyperplane~${x_a = x_b}$ defined by the inequalities~$x_{a'} < x_a$ for all~$a' \in A$ and~$x_b < x_{b'}$ for all~$b' \in B$.
%$\Sigma(a, b, A, B) \eqdef \set{\b{x} \in \R^n}{x_a = x_b \text{ and } x_{a'} < x_a \text{ for all } a' \in A \text{ and }  x_b < x_{b'} \text{ for all } b' \in B}$.
In other words, the shards decompose each hyperplane~$x_a = x_b$ of the braid arrangement into~$2^{b-a-1}$ cones.

The \defn{quotient fan} of a congruence~$\equiv$ of the weak Bruhat order is the polyhedral fan~$\c{F}_\equiv$~where
\begin{itemize}
\item the maximal cones are obtained by glueing together the chambers of the braid arrangement corresponding to permutations in the same congruence class of~$\equiv$,
\item the union of the codimension~$1$ cones is the union of the shards of the arcs of~$\c{A}_\equiv$.
\end{itemize}
(These two descriptions are equivalent.)
By construction, the braid fan refines the quotient fan~$\c{F}_\equiv$, and the dual graph of the quotient fan~$\c{F}_\equiv$ is isomorphic to the cover graph of the quotient~$\f{S}_n/{\equiv}$.
For instance, \cref{fig:fans}\,(right) represents the \defn{sylvester fan} (the quotient fan of the sylvester congruence of \cref{fig:sylvesterCongruence}\,(middle)) for~$n = 4$, whose dual graph is the cover graph of the Tamari lattice of \cref{fig:sylvesterCongruence}\,(right).

@@

\begin{figure}
	\capstart
	\centerline{\includegraphics[scale=.6]{permutahedronLeft4} \qquad \raisebox{-.4cm}{\includegraphics[scale=.6]{associahedron4}}}
	\caption{The permutahedron~$\Perm(4)$ (left) and the associahedron~$\Asso(4)$. \cite[Fig.~1]{PilaudSantos-quotientopes} \&~\cite[Fig.~5]{PadrolPilaudRitter}}
	\label{fig:quotientopes}
\end{figure}

A \defn{quotientope} for a lattice congruence~$\equiv$ of the weak Bruhat order is a polytope whose normal fan is the quotient fan~$\c{F}_\equiv$.
In particular, the skeleton of a quotientope is isomorphic to the cover graph of the quotient~$\f{S}_n/{\equiv}$.
For instance, the associahedron~$\Asso(n)$ (\cref{fig:quotientopes}\,(right) when~$n = 4$)  is a quotientope for the sylvester congruence (\cref{fig:sylvesterCongruence}\,(middle) when~$n = 4$), its normal fan is the sylvester fan (\cref{fig:fans}\,(right) when~$n = 4$), and its skeleton is the cover graph of the Tamari lattice (\cref{fig:sylvesterCongruence}\,(right) when~$n = 4$).
The existence of quotientopes was first proved in \cite{PilaudSantos} and later better understood in~\cite{PadrolPilaudRitter} using Minkowski sums of shard polytopes.

The \defn{Minkowski sum} of two polytopes~$\polytope{P}$ and~$\polytope{Q}$ is~$\polytope{P} + \polytope{Q} \eqdef \set{\b{p} + \b{q}}{\b{p} \in \polytope{P} \text{ and } \b{q} \in \polytope{Q}}$.
Recall that for any vector~$\b{v} \ne \b{0}$, the face of~$\polytope{P} + \polytope{Q}$ maximizing the scalar product with~$\b{v}$ is the Minkowski sum of the faces of~$\polytope{P}$ and~$\polytope{Q}$ maximizing the scalar product with~$\b{v}$.
In particular, if~$\b{v}$ is a generic direction, the vertex of~$\polytope{P} + \polytope{Q}$ extremal in direction~$\b{v}$ is just the sum of the vertices of~$\polytope{P}$ and~$\polytope{Q}$ that are extremal in direction~$\b{v}$.
Moreover, the normal fan of the Minkowski sum~$\polytope{P} + \polytope{Q}$ is the common refinement of the normal fans of~$\polytope{P}$ and~$\polytope{Q}$.

@@

\begin{figure}
	\capstart
	\centerline{\includegraphics[scale=1]{shardPolytopes}}
	\caption{The shard polytopes of the up arc~$(1, 4, \{2, 3\}, \varnothing)$, down arc~$(1, 4, \varnothing, \{2, 3\})$, yin arc~$(1, 4, \{2\}, \{3\})$, and yang arc~$(1, 4, \{3\}, \{2\})$. The vertices of the shard polytope of an arc~$\alpha \eqdef (a, b, A, B)$ are labeled by the corresponding $\alpha$-alternating matchings, where we use solid dots~$\bullet$ for elements in~$\{a\} \cup A$ and hollow dots~$\circ$ for elements in~$B \cup \{b\}$. The corresponding vertex coordinates are directly read replacing~$\bullet$ by~$1$ and~$\circ$ by~$-1$. For instance, the vertex labeled~${\bullet \cdot \cdot \,\circ}$ has coordinates~$(1,0,0,-1)$. Adapted from~\cite[Fig.~10]{PadrolPilaudRitter}.}
	\label{fig:shardPolytiopes}
\end{figure}

@@

\begin{lemma}
  \label{lem:submodsum}
  \vincent{I would not make this one so visible in a proper thm environment.}
  Consider a collection $\polytope{P}_1, \polytope{P}_2,\ldots ,\polytope{P}_k$ of deformed permutahedra defined by the submodular functions $f_1,f_2,\ldots ,f_k$. Then the Minkowski sum $\sum_i \polytope{P}_i$ is defined by the submodular function $f \eqdef \sum_i f_i$.
\end{lemma}

@@

The description of the weak rectangulotope~$\WRP(n)$ as a Minkowski sum in \cref{lem:weakMinkowski} allows us to easily compute the corresponding submodular function.
\vincent{Say here that all rays are preserved}
